From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-17 Pi+/paper/finite.tex
Outline from Jacek's text

Alternatively, groups can also be presented using generators and relations, or a quotient of the free group \ldots
\subsection{Permutations}~\label{subsec:permutations}

>>>>

Alternatively, groups can also be presented using generators and relations, or a quotient of the free group \ldots

2. In the usual setting, there are are many relations that are true for the elements of the group. For example, in the
(Z mod 3), we have a relation 1 + 1 + 1 = 0.

3. A group that has no other relations than the ones directly implied by
the definition of a group is called free.

4. A group is said to be generated by a set of its elemenst (e1, ..., en)
if every element in the group can be constructed from these elements
composing operations of + and inverse. For example, Z is a free group,
generated by one-element set (1).
4a. Thus, a free group g can be thought as sequences of letters from the
generating set, divided by the congruence closure of the group relation.
Definition: Free group.

5. One useful way of defining groups is by their presentations.
Presentation of a group builds a group by specifying what is the set of
generators - thus making a free group - a collection of equations that have
to be satisfied in the resulting group. For example, we could take the
generators to be the set (1) (giving out Z), and then add an equation (1 +
1 + 1 = 0) - thus, getting (Z mod 3).

5a. The generatos can be thought of as primitive operations in a language,
and the group structure gives us the way of composing these operations and
inverting them.

6. Building a group like that is possible because of the universal property
of the free group.
Theorem: Universal property of a free group.
Definition: formal definition of a free group.

7. Why do it like that? Because it is a very syntactic approach to doing
the theory. The way described previously operates directly in the semantics
of the group operation - equality of elements (such as the result of adding
two elements) has to be computed externally.
Here, the equality can be computed using a rewriting system specified by
the group axioms and the additional equations provided. However, because
equations are not directed, it is not always possible to construct a
well-behaving rewriting system. In general, the word problem (i.e. deciding
if two words are equal) is proven to be undecidable.

\subsection{Permutation Groups}

8. The group that we are interested in is the group of permutations on a
fininte set. We can define it formally in two different ways:
8a. Taking the type Aut (Fin n) and showing that it satisfies the group
axiom.
8b. Writing up a presentation - choosing generating set and equations that
have to be satisfied.

9. By proving these two to be equivalent, we are bridging the gap between
syntactic and semantic side in the whole proof - by assinging the meaning
(a bijection) to a word (a program).
9a. However, proving them equivalent will require, in particular, solving
the word problem (in this specific instance).

10. However, in the point 8b, there is an element of choice. One group can
be presented in many different ways. For example, we could generate the
permutations by using generators that:
 - swap i-th element with j-th element
 - swap i-th element with the first element
 - shift elements in a cyclic way: operation with label "i" leaves elements
with numbers > i in place, and element j <= i goes to (j + 1) mod i.

11. One way of thinking about these presentations is to compare them to
sorting algorithms with different primitive operations. If using a set of
operations, one can sort any list, then these operations generate the
permutations group. One weird example: pancake sorting.

\subsection{Coxeter Presentation}

12. Here, we are choosing a presentation based on the adjacent
transpositions generators - i.e. primitive operations of the form "swap
i-th and i+1 elements". In the language of sorting, this corresponds to
what the insertion sort or bubble sort are doing.
In the explanations below, we will use diagrams to draw the permutations.

12a. There two very obvious laws that such operation has to preserve:
 - Swapping the same two elements two times in a row does not change
anything (cancel)
TODO diagram.
 - When swapping two pairs of elements on a positions far away, it does not
matter in which order one does it (swap). (We can slide the wires).
TODO diagram.
But there is also one more law that these operations have to satisfy, in
the situation when we are swapping adjacent pairs of elements. It is best
seen on a diagram (braid)
TODO diagram.

This construction is called a Coxeter presentation. Writing it formally, we
get the following definition
Definition: Coxeter presentation.

13. There are two main considerations to this choice:
13a. How well the presentation matches the syntactic side of Pi? This is,
how closely does the operational semantics built from these particular
primitive operations match the natural operational semantics of Pi?
13b. How convenient is it to turn the presentation it into rewriting system
and solve the word problem for it?

Thus, the argument:
14. TODO I don't know what to write here, we have to think.

15a. The equations that define Coxeter presentation have a nice property:
as it is written, the LHS is smaller (lexicographically) than the RHS.
Thus, if we could direct them from right to left, it would have a
termination property out of the box.
15b. However, it is not at all obvious that the system would be confluent.
There are many possible critical pairs, such as
TODO diagram 1212 (braid, braid)
or
TODO diagram 3233 (braid, cancel)
As it turns out, there are instances of critical pairs that do not resolve.
A simplest example is the following:
TODO diagram 3213

16. The equations have to be changed. We thus propose a new presentation.
We keep the (cancel) and (swap) equations, but replace (braid) with a more
general:
TODO diagram:

Definition: Long Coxeter presentation.

17. In return for increased complexity of the equations, it satisfies the
diamond property (if we direct the equation). Thus, together with the
(preserved) termination, it creates a confluent rewriting system.
Proof: All the critical pairs resolve, in the appendix.
Definition: A normal form is a list that cannot be reduced any further.

18. We define the equivalence classes on List (Fin n) to be precisely these
elements that have the same normal form.

19. However, the Long Coxeter relations are unwieldly and have no
connectino to the broader scope of computational group theory - while the
usual Coxeter presentation is to-go choice in this area (cite all these
papers in Coq, this guy from Sweden(?) etc.).
Thus, we prove:
Proposition: Long Coxeter is equivalent to normal Coxeter.

19a. (Note: here we try to convince the reader about reusability badge) So,
the Long Coxeter can be regarded as a trick to get the proof to pass - but
the external user-facing interface is still that of the usual Coxeter
presentation.

20. Now, to prove the equivalence of 9, on one side we have Aut(Fin n), and
on the other, equivalence classes of List (Fin n) wrt to the usual Coxeter
group, we would define two functions from and back. The easiest way to
define a function out of the group presentation is to define them on the
equivalence classes' representatives. But to do that, we have to know how
do these representatives look like exactly.

\subsection{Lehmer Codes}

21. Enter: Lehmer codes.
TODO diagram with List (Fin n) being divided into equivalence classes, and
Lehmer codes being an image of immersion, being the normal form
(representative) in each class.

\subsection{Permutations}~\label{subsec:permutations}
